---
title: "GISP all drugs"
author: "Scott Olesen"
output: pdf_document
---

```{r global_options, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  cache = TRUE, autodep = TRUE,
  fig.path = 'fig/', dev = c('pdf', 'png')
)

library(dplyr)
library(readr) # for read_tsv
library(ggplot2)
library(stats) # for nls
library(knitr) # for making the R markdown doc
library(forcats) # for fct_relevel
library(purrr) # for map, map_dbl, etc.
```

# Antibiotic use

```{r load_use_data}
denom = read_tsv('../data/denom.tsv')

counts = read_tsv('../data/abx_monthly_counts.tsv') %>%
  # make Jan = 0
  mutate(month = mon - 1) %>%
  select(-mon)

use_data = left_join(counts, denom, by = c('state', 'year')) %>%
  # group by Census region
  mutate(region = state.region[match(state, state.name)]) %>%
  group_by(year, month, region, antibiotic) %>%
  summarize_at(c('n_rx', 'n_member'), sum) %>%
  ungroup() %>%
  # claims per 1k people per year
  mutate(ckpy = n_rx / n_member * 1000)
```

I group drugs into classes:

- macolides (azi, clarithro, erythro)
- quinolones (*xacin)
- celaphosporins (cef-, ceph-)
- tetracyclines (*cycline)
- penicillins (amox, amox/clav, pen v, amp, diclox, pen g)

```{r drug_class_seasonality}
use_by_class = use_data %>%
  mutate(class = case_when(
    .$antibiotic %in% c('azithromycin', 'clarithromycin', 'erythromycin') ~ 'macrolide',
    .$antibiotic %in% c('amoxicillin', 'amoxicillin/clavulanate', 'penicillin v', 'ampicillin', 'dicloxacillin', 'penicillin g') ~ 'penicillin',
    str_detect(.$antibiotic, 'cycline$') ~ 'tetracycline',
    str_detect(.$antibiotic, 'xacin$') ~ 'quinolone',
    str_detect(.$antibiotic, '^ce[f|(ph)]') ~ 'cephalosporin',
    TRUE ~ 'other'
  )) %>%
  filter(class != 'other') %>%
  group_by(region, year, month, class) %>%
  summarize_at('ckpy', sum) %>%
  ungroup() %>%
  mutate(
    year_month = year + month / 12,
    t = (year - min(year)) * 12 + month
  )
```

By eye, it looks like there are seasonal patterns:

```{r use_by_class_plot}
use_by_class %>%
  ggplot(aes(year_month, ckpy)) +
  facet_wrap(~ region) +
  geom_line(aes(color = class)) +
  xlab('')
```

But there is also something fishy about the 2014-2015 merged: use seems to go up for all drugs, especially in the West. This might introduce some weird effects in the wavelet analysis.

## Wavelet analysis

To detect seasonality of antibiotic use, I do a wavelet analysis. I exclude the 2015 data because of the weird 2014-2015 jump, which introduces some weird stuff into the wavelets.

```{r compute_use_wavelets, results = 'hide'}
fence = function(x, lower, upper) pmax(lower, pmin(upper, x))

wt_plot_data = function(w) {
  with(w, {
    crossing(i = seq_along(axis.1), j = seq_along(axis.2)) %>%
      mutate(x = axis.1[i], y = axis.2[j],
             scale = Scale[j],
             amplitude = map2_dbl(j, i, ~ Ampl[.x, .y]),
             power = map2_dbl(j, i, ~ Power[.x, .y]),
             p.value = map2_dbl(j, i, ~ Power.pval[.x, .y]),
             coi_raw = approx(coi.1, coi.2, x)$y,
             coi = fence(coi_raw, min(y), max(y)),
             sig = as.double(p.value < 0.05))
  })
}

wavelet_region_class = use_by_class %>%
  filter(year < 2015) %>%
  nest(-region, -class) %>%
  mutate(
    wavelet = map(data, ~ WaveletComp::analyze.wavelet(., 'ckpy', verbose = FALSE)),
    wavelet_plot_data = map(wavelet, wt_plot_data)
  )
```

Dotted lines show 12 month periods. (The Northeast/tetracyclines has an extra line for 6 month period.) Solid black lines show regions with $p < 0.05$. Shaded areas along the side are the "cone of influence" where edge effects are important.

```{r show_use_wavelets}
# this is a kludge to make the x-axis look right. it should be drawn more directly
# from the data.
max_x = 47

y_1yr = wavelet_region_class$wavelet_plot_data[[1]] %>%
  mutate(d = abs(scale - 12)) %>%
  filter(d == min(d)) %$%
  unique(y)

y_6mo = wavelet_region_class$wavelet_plot_data[[1]] %>%
  mutate(d = abs(scale - 6)) %>%
  filter(d == min(d)) %$%
  unique(y)

# make a dummy dataset
ne_tet = data_frame(
  region = 'Northeast', class = 'tetracycline',
  x = 0, xend = max_x,
  y = y_6mo, yend = y)

wavelet_region_class %>%
  select(region, class, wavelet_plot_data) %>%
  unnest %>%
  ggplot(aes(x - 1, y)) +
  geom_tile(aes(fill = amplitude)) +
  geom_ribbon(aes(ymin = coi, ymax = max(y)), alpha = 0.25) +
  geom_contour(aes(z = as.double(p.value < 0.05)), color = 'black', size = 0.5, bins = 1) +
  facet_grid(class ~ region) +
  scale_fill_distiller(palette = 'RdYlBu') +
  geom_segment(x = 1, xend = max_x + 1, y = 3.65, yend = 3.65, linetype = 2) +
  geom_segment(data = ne_tet, aes(xend = xend - 1, y = y, yend = yend), linetype = 2) +
  xlab('time (months since Jan 2005)') +
  ylab('log2 period (months)') +
  theme_minimal()
```

- In general, macrolides, penicillins, and quinolones are highly seasonal.
- Cephalosporins are seasonal in the South and North Central (i.e., Midwest) and much less so in the Northeast and West.
- Tetracyclines are weakly seasonal in North Central and West. In the Northeast, there is a 6-monthly seasonality.

## Sinusoidal fits

```{r use_data}
omega = 2 * pi / 12

use_nls_f = function(df) {
  nls(ckpy ~ amplitude * sin(omega * (t - phase)) + offset,
      start = list(amplitude = 1.0, phase = 0.0, offset = 10.0),
      data = df)
}

peak_time = function(phase, amplitude) {
  phase %>%
    # add 3 because a sine's peak is 3 months after it's phase
    { . + 3 } %>%
    # if amplitude is negative, add another 6 months to flip the period
    { if_else(amplitude < 0, . + 6, .) } %>%
    # subtract whole years until you get to 0-12
    { . %% 12 } %>%
    # make it between -6 and +6
    { if_else(. > 6, . - 12, .) }
}

use_by_class %>%
  nest(-region, -class) %>%
  mutate(
    nls_model = map(data, use_nls_f),
    raw_phase = map_dbl(nls_model, ~ coef(.)['phase']),
    raw_amplitude = map_dbl(nls_model, ~ coef(.)['amplitude']),
    amplitude_pval = map_dbl(nls_model, ~ coef(summary(.))['amplitude', 'Pr(>|t|)']),
    amplitude_sig = p.adjust(amplitude_pval, 'BH') < 0.05,
    peak = map2_dbl(raw_phase, raw_amplitude, peak_time),
    amplitude = abs(raw_amplitude)
  ) %>%
  mutate(class = fct_reorder(class, -amplitude)) %>%
  select(class, region, amplitude, amplitude_sig, peak) %>%
  arrange(class, region) %>%
  mutate_if(is.numeric, ~ round(., 3)) %>%
  kable(caption = 'Sinusoidal fits to use data')
```

Use of every drug is yearly-seasonal, except for cephalosporins in the West and tetracyclines in the Northeast. I expect the Northeast tet thing has something to do with the 6-month cycle.

### Tetracyclines

When looking just at the tetracycline drugs, it's clear that minocycline (used to treat acne) is stable through the year, while doxycycline (used to treat Lyme disease, among other things) has more complicated seasonal dynamics:

```{r tetracyclines}
use_data %>%
  filter(str_detect(antibiotic, 'cycline$')) %>%
  mutate(year_month = year + month / 12) %>%
  ggplot(aes(year_month, ckpy)) +
  facet_wrap(~ region) +
  geom_line(aes(color = antibiotic))
```

# Resistance

```{r load_res_data}
clinics = read_tsv('../db/gisp_sites.tsv') %>%
  rename(clinic = abbreviation, state_abbreviation = state) %>%
  mutate(state_id = match(state_abbreviation, state.abb),
         region = state.region[state_id]) %>%
  select(clinic, region)

gisp = read_tsv('../../gisp/data/gisp.tsv') %>%
  left_join(clinics, by = 'clinic') %>%
  select(-week) %>%
  gather('drug', 'mic', penicillin:azithromycin) %>%
  filter(!is.na(mic)) %>%
  # Jan -> 0, Feb -> 1, etc.
  mutate(month = match(month, month.abb) - 1,
         t = (year - min(year)) * 12 + month,
         dil = round(log2(mic)),
         clinic_year = str_c(clinic, '_', year)) %>%
  group_by(clinic_year, drug) %>%
  filter(n() >= 10) %>%
  ungroup()

# assert all months are between 0 and 11
stopifnot(all(between(gisp$month, 0, 11)))
```

## Model fit

This is where the model fits would go for the drugs beyond macrolides. The code is in the block below. The fits are fairly slow, so it's nice for development purposes to have some way to save the fit data and re-load it.

```{r fit_resistance, eval = FALSE}
nls_res_fit = function(df) {
  n_cys = length(unique(df$clinic_year))

  model_matrix = model.matrix(~ 0 + clinic_year, df) %>%
    set_colnames(str_replace(colnames(.), '^clinic_year', ''))

  stopifnot(dim(model_matrix) == c(nrow(df), n_cys))

  # use clinic/year means as best start guess
  start_intercepts = df %>%
    group_by(clinic_year) %>%
    summarize(value = mean(dil)) %>%
    arrange(clinic_year) %T>%
    # check that the order of these values matches the model matrix columns
    { stopifnot(all(.$clinic_year == colnames(model_matrix))) } %>%
    pull(value)
  
  # as a first guess for clinic/year slopes, use 0
  start_slopes = rep(0, n_cys)
  
  start = list(A = 0.05, phase = 0.0, intercept = start_intercepts, slope = start_slopes)

  # sinusoidal + linear fit function
  fit_f = function(month, A, phase, slope, intercept) {
    intercept_term = drop(model_matrix %*% intercept)
    slope_term = drop(model_matrix %*% slope) * month
    A * sin(2 * pi / 12 * (month - phase)) + slope_term + intercept_term
  }

  # fit the model
  model = nls(dil ~ fit_f(month, A, phase, slope, intercept),
              start = start,
              data = df)
  
  model
}
```

It might make sense to do models for each drug/region combination, since for some drugs it seems like use patterns vary by region.
